---
title: "Analysing Long Read RNA-Seq data with bambu"
author: Chen Ying, Yuk Kei Wan, Jonathan Göke
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bambu}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning=FALSE, message=FALSE,
  comment = "##>"
)
```


## Introduction
*[Bambu](https://github.com/GoekeLab/bambu)* can be used for transcript discovery and quantification from long read RNA-Seq data. Here, we present an example workflow for analysing Nanopore long read RNA-Sequencing data from 2 human cancer cell lines


### Content
* [Quick Start: Transcript discovery and quantification with bambu](#quick-start)
* [Input data](#input-data)
  + [Aligned reads (bam files)](#bam-files)
  + [Genome sequence (fasta file/ BSGenome object)](#genome-sequences)
  + [Genome annotations (bambu annotations object/ gtf file / TxDb object)](#annotations)
* [Transcript discovery and quantification](#transcript-discovery-quantification)
  + [Run bambu](#run-bambu)
  + [Visualise results](#visualise-results)
  + [Obtain gene expression estimates from transcript expression](#gene-expression)
  + [Save data (gtf/text)](#save-data)
* [Identify differentially expressed genes](#DESeq2)
* [Identify differential transcript usage](#DEXSeq)
* [Running bambu with large number of samples](#large-sample-num)

## Data description {#quick-start}
To demonstrate the usage of *Bambu*, we used long-read RNA-Seq data generated using Oxford Nanopore Sequencing from the NanoporeRNASeq package, which consists of 6 samples from two human cell lines (K562 and MCF7). Each of these cell lines has three replicates, with 1 direct RNA sequencing data and 2 cDNA sequencing data. Reads are aligned to chromosome 22 (Grch38) and stored as bam files.

```{r, include = FALSE}
devtools::install_github("GoekeLab/NanoporeRNASeq")
```

```{r setup}
library(NanoporeRNASeq)
library(bambu)
```

## Input data {#input-data}
load sample information
```{r}
data("sample_info")
sample_info
```

### Load Bam files {#bam-files}
```{r}
data("bamFiles")
```

### Load genome sequences {#genome-sequences}
```{r}
# get path to fasta file
fasta_path <- system.file("extdata", "Homo_sapiens.GRCh38.dna_sm.primary_assembly_chr22_1_25409234.fa", package="NanoporeRNASeq")
# BSgenome
library(library(BSgenome.Hsapiens.NCBI.GRCh38))
```

### Load Annotation {#annotation}
```{r}
data("annotation")
```

## Transcript discovery and quantification {#transcript-discovery-quantification}
### Apply *bambu* to perform EM on  extended annotations {#run-bambu}
```{r}
se <- bambu(reads = bamFiles, annotations = annotation, genomeSequence = "BSgenome.Hsapiens.NCBI.GRCh38", extendAnnotations = TRUE)
se
```
*bambu* allows quantification without isoform discovery
```{r}
seUnextended <- bambu(reads = bamFiles, annotations = annotation, genomeSequence = "BSgenome.Hsapiens.NCBI.GRCh38")
seUnextended
```

### Check expression estimates {#visualise-results}
We can check the estimated transcript expression using heatmap:
```{r, fig.width = 8, fig.height = 6}
colData(se)$groupVar <- sample_info$cellLine
colnames(se) <- sample_info$sample_id
colData(se)$name <- sample_info$sample_id
plot.bambu(se, group.variable = "groupVar", type = "heatmap")
```

or with PCA plot
```{r, fig.width = 8, fig.height = 6}
colData(se)$groupVar <-  sample_info$cellLine
plot.bambu(se, group.variable = "groupVar", type = "pca")
```


### Check for gene examples 

Single gene examples can also be checked using plot functions from *bambu*
```{r, fig.width = 8, fig.height = 10}
plot.bambu(se, type = "annotation", gene_id = unique(rowData(se)$GENEID)[10])

```

### Transcript to Gene expression {#gene-expression}
Gene expression can be obtained from transcript expression using this function:
```{r}
seGene <- transcriptToGeneExpression(se)
seGene
```

Gene expression heatmap
```{r, fig.width = 8, fig.height = 6}
colData(seGene)$groupVar <-  sample_info$cellLine
plot.bambu(seGene, group.variable = "groupVar", type = "heatmap")
```

Gene expression PCA plot
```{r, fig.width = 8, fig.height = 6}
colData(seGene)$groupVar <-  sample_info$cellLine
plot.bambu(seGene, group.variable = "groupVar", type = "pca")
```

### Save data (gtf/text){#save-data}
```{r}
writeBambuOutput(se, path = "./bambu/")
```

## Differentially expressed genes {#DESeq2}
We used *DESeq2* to find the differentially expressed genes:
```{r}
library(DESeq2)
dds <- DESeqDataSetFromMatrix(apply(assays(seGene)$counts,c(1,2),round),#tmp_wide[,-1],
                                   colData = colData(se),
                                   design = ~ groupVar)
system.time(dds.deseq <- DESeq(dds))

deGeneRes <- DESeq2::results(dds.deseq, independentFiltering=FALSE)
```

```{r}
head(deGeneRes[order(deGeneRes$padj),])
```

```{r}
summary(deGeneRes)
```
```{r, fig.width = 8, fig.height = 6}
plotMA(deGeneRes, ylim = c(-3,3))
```

Plotting shrinked lFC results
```{r, fig.width = 8, fig.height = 6}
library(apeglm)
resLFC <- lfcShrink(dds.deseq, coef="groupVar_MCF7_vs_K562", type="apeglm")
plotMA(resLFC, ylim = c(-3,3))
```

## Differential expression for isoform detection {#DEXSeq}
We used *DEXSeq* to detect alternative used isoforms.
```{r}
library(DRIMSeq)
count.data <- as.data.frame(rowData(se))
count.data$gene_id <- count.data$GENEID
count.data$feature_id <- count.data$TXNAME
count.data$GENEID <- count.data$TXNAME <- NULL

count.data <- cbind(count.data, assays(se)$counts)

sample.info <- as.data.frame(colData(se))
sample.info$sample_id <- sample.info$name
sample.info$name <- NULL
d <- dmDSdata(counts=count.data, samples=sample.info)

n_samp_gene <- 1
n_samp_feature <- 1
min_count_gene <- 1
min_count_feature <- 1
dFilter <- dmFilter(d,
              min_samps_feature_expr = n_samp_feature, 
              min_samps_feature_prop = n_samp_feature,
              min_samps_gene_expr = n_samp_gene, 
              min_feature_expr = min_count_feature,
              min_gene_expr = min_count_gene,
              min_feature_prop=0.1)
table(table(counts(dFilter)$gene_id)) ## number of isoforms 



```

```{r}
library(DEXSeq)
formulaFullModel <- as.formula("~sample + exon + groupVar:exon")


dxd <- DEXSeqDataSet(countData=round(as.matrix(counts(dFilter)[,-c(1:2)])),
                     sampleData=DRIMSeq::samples(dFilter),
                     design=formulaFullModel,
                     featureID = counts(dFilter)$feature_id,
                     groupID=counts(dFilter)$gene_id)


system.time({
  dxd <- estimateSizeFactors(dxd)
  print('Size factor estimated')
  dxd <- estimateDispersions(dxd, formula = formulaFullModel)
  print('Dispersion estimated')
  #dxd <- estimateExonFoldChanges( dxd )
  dxd <- testForDEU(dxd, fullModel = formulaFullModel)
  print('DEU tested')
  dxd <- estimateExonFoldChanges(dxd, fitExpToVar="groupVar")
  print('Exon fold changes estimated')
}) 
```

```{r}
dxr <- DEXSeqResults(dxd, independentFiltering=FALSE)
head(dxr)
```


```{r}
library(stageR)
strp <- function(x) substr(x,1,15)
qval <- perGeneQValue(dxr)
dxr.g <- data.frame(gene=names(qval),qval)

columns <- c("featureID","groupID","pvalue")
dxr_pval <- as.data.frame(dxr[,columns])
head(dxr_pval)

pConfirmation <- matrix(dxr_pval$pvalue,ncol=1)
dimnames(pConfirmation) <- list(strp(dxr_pval$featureID),"transcript")
pScreen <- qval
names(pScreen) <- strp(names(pScreen))
tx2gene <- as.data.frame(dxr_pval[,c("featureID", "groupID")])
for (i in 1:2) tx2gene[,i] <- strp(tx2gene[,i])


stageRObj <- stageRTx(pScreen=pScreen, pConfirmation=pConfirmation,
                      pScreenAdjusted=TRUE, tx2gene=tx2gene)
stageRObj <- stageWiseAdjustment(stageRObj, method="dtu", alpha=0.5)
suppressWarnings({
  dex.padj <- getAdjustedPValues(stageRObj, order=FALSE,
                                 onlySignificantGenes=TRUE)
})
```

```{r}
dxrDT <- data.table(as.data.frame(dxr))
setnames(dxrDT, old = c('groupID','featureID'), new = c('geneID','txID'))
dex.padj <- data.table(dex.padj)

dxrDT <- dex.padj[dxrDT, on = c('geneID','txID')]
head(dxrDT)
```


```{r, fig.width = 8, fig.height = 6}
dxrDT[,sigLFC2:=(padj < 0.5&(abs(log2fold_MCF7_K562)>=2))]
ggplot(dxrDT, aes(y = log2fold_MCF7_K562, x = exonBaseMean, color = as.factor(padj<0.5)))+
  geom_point(size = 0.5)+
     scale_x_log10()+
  scale_color_manual(values = c('grey','indianred'), name = "Significant")+
     xlab("Mean of normalized counts")+
    ylab("Log2 Fold change")+
  theme_minimal()

```

```{r}
dxrDT[padj<0.5,.(geneID, txID, log2fold_MCF7_K562,K562,MCF7)]
```
### Running bambu with large number of samples {#large-sample-num}


```{r}
sessionInfo()
```



